---
title: "Sample ICU Study Testing"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting Up the Test Environment

Before running the tests, ensure that you have installed the necessary packages. You can install them using the following command:

```
install.packages(c("dplyr", "CDMConnector", "omopgenerics", "testthat", "PatientProfiles", "here"))
```

## Running the Tests with `testthat`

The following code executes three separate tests:

- **Checking cohort counts**
- **Characterization**
- **Running the study**

## Creating a Test Database Using Excel

- This project includes a sample population dataset, `icu_sample_population`, located in the `tests/testthat` folder. It contains:
  - 5 patients with ICU visits
  - 7 patients with COVID-19
  - 5 patients with mechanical ventilation

  Explore the dataset and modify the tables to create your own test cases.

- The `TestGenerator` package provides a blank CDM, allowing you to input your own patient data from scratch.

- You can also create a test database for one of your ongoing studies.

- If needed, you can collaborate with a data scientist, or we can assist in processing the patients and running the tests for you.

```{r}
cdmVersion <- "5.3"

TestGenerator::generateTestTables(
  tableNames = c("person", "drug_exposure", "condition_occurrence"),
  cdmVersion = cdmVersion,
  outputFolder = here::here("tests", "testthat"),
  filename = paste0("test_cdm_", cdmVersion)
)
```

## Reading the Test Data

After generating the test patients, a data scientist can use the Excel file to create the test CDM.

```{r test_data}
filePath <- testthat::test_path("icu_sample_population.xlsx")
TestGenerator::readPatients.xl(
  filePath = filePath,
  testName = "icu_sample_population",
  cdmVersion = "5.3"
)
```

## Creating the CDM Database in Memory

Now, we can select a dataset to create our CDM reference.

```{r}
cdm <- TestGenerator::patientsCDM(testName = "icu_sample_population")
```

## Exploring the Data as a Standard CDM Reference

The CDM reference contains all tables following standard conventions.

```{r}
cdm

cdm$person
```

## Creating Cohorts and Running Tests in the `testthat` Environment

The following steps are performed:

- Checking cohort counts
- Testing the characterization function
- Running the study

```{r, eval=FALSE}
library(dplyr)
library(CDMConnector)
library(omopgenerics)
library(testthat)
library(PatientProfiles)
library(here)
source(here::here("R/icu.R"))

test_that("Check cohort counts for COVID, ICU visits, and ventilation", {
  cdm <- TestGenerator::patientsCDM(testName = "icu_sample_population")
  
  # Generate cohorts
  cohorts <- here::here("cohorts", "population")
  icu_cohort_sets <- CDMConnector::readCohortSet(cohorts)
  cdm <- CDMConnector::generate_cohort_set(cdm, icu_cohort_sets, name = "icu_population_cohorts")
  cdm[["icu_population_cohorts"]] %>% omopgenerics::settings()
  
  # Check counts
  cohort_counts <- omopgenerics::cohortCount(cdm$icu_population_cohorts)
  
  expect_equal(cohort_counts[1, ] %>% pull(number_records), 5)
  expect_equal(cohort_counts[2, ] %>% pull(number_records), 7)
  expect_equal(cohort_counts[3, ] %>% pull(number_records), 5)
})

test_that("Characterization", {
  cdm <- TestGenerator::patientsCDM(testName = "icu_sample_population")
  cohorts <- here::here("cohorts", "population")
  
  icu_cohort <- CDMConnector::readCohortSet(cohorts) %>%
    dplyr::filter(cohort_name == "p2_c1_014_icu_visit_dv_v2")
  cdm <- CDMConnector::generate_cohort_set(cdm, icu_cohort, name = "icu")
  
  covid_cohort <- CDMConnector::readCohortSet(cohorts) %>%
    dplyr::filter(cohort_name == "p2_c1_014_covid19_v2")
  cdm <- CDMConnector::generate_cohort_set(cdm, covid_cohort, name = "covid")
  
  ventilation_cohort <- CDMConnector::readCohortSet(cohorts) %>%
    dplyr::filter(cohort_name == "p2_c1_014_mechanical_ventilation_dv_v2")
  cdm <- CDMConnector::generate_cohort_set(cdm, ventilation_cohort, name = "ventilation")
  
  summaryCharacterization <- characterizeICU(
    cdm = cdm,
    icuCohort = "icu",
    covidCohort = "covid",
    ventilationCohort = "ventilation"
  )
  
  expect_s3_class(summaryCharacterization, c("summarised_result", "omop_result"))
  expect_equal(
    summaryCharacterization %>% 
      filter(variable_name == "covid_m14_to_0", estimate_name == "count") %>% 
      pull(estimate_value), "4"
  )
  expect_equal(
    summaryCharacterization %>% 
      filter(variable_name == "ventilation_m7_to_0", estimate_name == "count") %>% 
      pull(estimate_value), "3"
  )
})

test_that("Run Study", {
  cdm <- TestGenerator::patientsCDM(testName = "icu_sample_population")
  outputDir <- paste0(tempdir(), "/results")
  dir.create(outputDir)
  
  runStudy(cdm, outputDir = outputDir)
  
  files_created <- list.files(outputDir)
  expect_true("icu_summary.csv" %in% files_created)
  
  unlink(outputDir, recursive = TRUE)
})
