---
title: "TestGenerator Workshop"
author: "Erasmus MC"
date: "2025-01-30"
output:
  slidy_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What is TestGenerator?

TestGenerator is an R package designed to create an OMOP-CDM test database with a sample of patients.

- Follows OMOP-CDM conventions
- Uses a complete vocabulary
- Ensures confidence that our code executes correctly

## Why it is a good idea to test study code?

- Verify that we are using the correct parameters in analytical packages
- Test custom code
- Ensure our study code runs successfully from start to finish

## How does it work?
### TestGenerator::readPatients.xl()

1. We write an Excel file, where: 
  - Sheets represents tables in the OMOP-CDM 
  - Rows represent records of patients
  
A familiar interface for non-programmers

## TestGenerator::readPatients.csv()

- Useful to load a test data such as MIMIC IV, NCR, etc. 
- We can replace vocabulary tables with a custom vocabulary

```{r}

list.files(testthat::test_path("ncr_test_cdm_10"))

```

## TestGenerator::readPatients.csv()

- Creates a test database with a sample of patients in json format. 
  - We can reuse this data to test our study code
  - Save multiple datasets if required for specific cases
  
````
{
    "cdm.person": [
        {
            "person_id":1,
            "gender_concept_id":8532,
            "year_of_birth":1999,
            "race_concept_id":0,
            "ethnicity_concept_id":0
        },
        {
            "person_id":2,
            "gender_concept_id":8507,
            "year_of_birth":1995,
            "race_concept_id":0,
            "ethnicity_concept_id":0
        }
    ],
    "cdm.observation_period" : [
        {
          "observation_period_id": 1,
          "person_id":1,
          "observation_period_start_date":"2000-01-01",
          "observation_period_end_date":"2001-01-01",
          "period_type_concept_id": 0
        },
        {
          "observation_period_id": 2,
          "person_id":2,
          "observation_period_start_date":"2000-01-01",
          "observation_period_end_date":"2001-01-01",
          "period_type_concept_id": 0
        }
    ],
    "cdm.drug_exposure": [
        {
          "drug_exposure_id": 1,
          "person_id":1,
          "drug_concept_id":40852223,
          "quantity": 30,
          "days_supply": 30,
          "drug_exposure_start_date":"2000-01-01",
          "drug_exposure_end_date":"2000-01-31",
          "drug_type_concept_id":0
        },
        {
          "drug_exposure_id": 2,
          "person_id":2,
          "drug_concept_id":40852223,
          "quantity": 30,
          "days_supply": 30,
          "drug_exposure_start_date":"2000-01-01",
          "drug_exposure_end_date":"2000-01-31",
          "drug_type_concept_id":0
        }
    ]
````

## TestGenerator::readPatients()

- Save multiple datasets if required for specific cases

- A directory with XLSX files

```{r, echo=TRUE}
list.files(testthat::test_path(), pattern = ".xlsx")
```

## TestGenerator::readPatients()

- Save datasets in JSON format.
  - This allows to save several datasets for testing
  
```{r, echo=FALSE}
mimicIVsubsetLocation <- here::here("tests", "testthat", "mimicIVsubset")
testCasesLocation <- here::here("tests", "testthat", "testCases")
icu_sample_population_location <- here::here("tests", "testthat", "icu_sample_population.xlsx")

```
  
```{r, echo=TRUE}
# This is an example of a test that uses the MIMIC IV subset dataset
TestGenerator::readPatients.csv(filePath = testthat::test_path("mimicIVsubset"),
                                testName = "mimicIVsubset")
```

## TestGenerator::readPatients()

```{r, echo=TRUE}
TestGenerator::readPatients.xl(filePath = icu_sample_population_location,
                               testName = "icu_sample_population")
```

## TestGenerator::readPatients()

```{r, echo=TRUE}
list.files(testCasesLocation)
```


## TestGenerator::readPatients()

```{r, echo=TRUE}
list.files(testCasesLocation)
```

## TestGenerator::patientsCDM()

```{r, echo=TRUE}
cdm <- TestGenerator::patientsCDM(pathJson = testCasesLocation,
                                  testName = "icu_sample_population")
```

## TestGenerator::patientsCDM()

```{r, echo=TRUE}
cdm
```


## TestGenerator::patientsCDM()

```{r, echo=TRUE}
cdm$person
```

## TestGenerator::patientsCDM()

```{r, echo=TRUE}
cdm$drug_exposure
```
## Test Scenarios

- Cohorts/Phenotyping 
- Study objectives
- Study execution (runStudy)

## Testing study P2-C1-014- ICU Prescriptions

- Cohorts

````
test_that("condition and visit cohorts counts", {
  # Read patients from JSON (run only once)
  # filePath <- testthat::test_path("icu_sample_population.xlsx")
  # outputPath <- file.path(tempdir(), "testCases")
  # dir.create(outputPath)
  # TestGenerator::readPatients.xl(filePath = filePath,
  #                                testName = "icu_sample_population",
  #                                outputPath = outputPath,
  #                                cdmVersion = "5.3")

  cdm <- TestGenerator::patientsCDM(testName = "icu_sample_population")
  # Cohorts
  population <- system.file("cohorts",
                            "population",
                            package = "P2C1014PrescriptionsICU")
  cohort_set <- CDMConnector::readCohortSet(population)

  cdm <- CDMConnector::generate_cohort_set(cdm,
                                           cohort_set,
                                           name = "icu_cohorts")

  # Counts

  cdm$icu_cohorts %>% collect() %>% filter(cohort_definition_id == 5)

  cohort_count <- omopgenerics::cohortCount(cdm$icu_cohorts) %>%
    dplyr::left_join(cohort_set, by = "cohort_definition_id") %>%
    dplyr::select(cohort_definition_id,
                  cohort_name,
                  number_records,
                  number_subjects)


  hosp_count <- cohort_count %>% filter(cohort_name == "p2_c1_014_hospitalization_dv_v2") %>% pull(number_subjects)
  hosp_icu_visit_count <- cohort_count %>% filter(cohort_name == "p2_c1_014_hospitalization_icu_visit_dv_v2") %>% pull(number_subjects)
  hosp_no_icu_count <- cohort_count %>% filter(cohort_name == "p2_c1_014_hospitalization_no_icu_visit_dv_v4") %>% pull(number_subjects)
  icu_only_count <- cohort_count %>% filter(cohort_name == "p2_c1_014_icu_visit_dv_v2") %>% pull(number_subjects)
  covid_count <- cohort_count %>% filter(cohort_name == "p2_c1_014_covid19_v2") %>% pull(number_subjects)
  mech_count <- cohort_count %>% filter(cohort_name == "p2_c1_014_mechanical_ventilation_dv_v2") %>% pull(number_subjects)

  # Test
  expect_equal(hosp_count, 8)
  expect_equal(hosp_icu_visit_count, 7)
  expect_equal(hosp_no_icu_count, 1)
  expect_equal(icu_only_count, 7)
  expect_equal(covid_count, 5)
  expect_equal(mech_count, 5)
  # unlink(outputPath, recursive = TRUE)

})

````

## Testing study P3-C1-015 - Incidence of myoclonus in heart failure

- Study objectives

- Treatment inclusion

````
test_that("runStudy results, treatment inclusion", {

  testPath <- getwd()
  # Read patients from JSON
  cdm <- TestGenerator::patientsCDM(
    pathJson = testPath,
    testName = "testTreatmentInclusion"
  )

  dbName <- "TestDataTreatmentInclusion"

  resultsDir <- file.path(testPath, glue::glue("Results_{dbName}"))

  runStudy(
    conn = attr(cdm, "dbcon"),
    cdmDatabaseSchema = "main",
    resultsDatabaseSchema = "main",
    dbName = dbName,
    tablePrefix = "pre_",
    minCellCount = 0
  )

  # define result file paths
  resultPathAceInhibitors <- file.path(resultsDir, "incidence_hf_treatment_ace_inhibitors.csv")
  resultPathArbs <- file.path(resultsDir, "incidence_hf_treatment_arbs.csv")
  resultPathSacubitrilFirst <- file.path(resultsDir, "incidence_hf_treatment_sacubitril_first.csv")
  resultPathSacubitrilLater <- file.path(resultsDir, "incidence_hf_treatment_sacubitril_later.csv")

  # load and format results
  resultAceInhibitors <- omopgenerics::importSummarisedResult(resultPathAceInhibitors)
  resultAceInhibitors <- formatIncidence(resultAceInhibitors)
  resultArbs <- omopgenerics::importSummarisedResult(resultPathArbs)
  resultArbs <- formatIncidence(resultArbs)
  resultSacubitrilFirst <- omopgenerics::importSummarisedResult(resultPathSacubitrilFirst)
  resultSacubitrilFirst <- formatIncidence(resultSacubitrilFirst)
  resultSacubitrilLater <- omopgenerics::importSummarisedResult(resultPathSacubitrilLater)
  resultSacubitrilLater <- formatIncidence(resultSacubitrilLater)


  # get treatment for female subjects
  resultAceInhibitorsFemale <- minimiseResults(resultAceInhibitors,
    gender = "Female",
    timeAtRisk = "0 to 30"
  )

  resultArbsFemale <- minimiseResults(resultArbs,
    gender = "Female",
    timeAtRisk = "0 to 30"
  )

  resultSacubitrilFirstFemale <- minimiseResults(resultSacubitrilFirst,
    gender = "Female",
    timeAtRisk = "0 to 30"
  )

  resultSacubitrilFirstLaterFemale <- minimiseResults(resultSacubitrilLater,
    gender = "Female",
    timeAtRisk = "0 to 30"
  )

  expect_equal(nrow(resultAceInhibitorsFemale)[[1]], 1)
  expect_equal(resultAceInhibitorsFemale$n_persons[1], 1)
  expect_equal(resultAceInhibitorsFemale$n_events[1], 1)
  expect_equal(resultAceInhibitorsFemale$incidence_start_date[1], "2020-07-02")

  expect_equal(nrow(resultArbsFemale)[[1]], 1)
  expect_equal(resultArbsFemale$n_persons[1], 1)
  expect_equal(resultArbsFemale$n_events[1], 1)
  expect_equal(resultArbsFemale$incidence_start_date[1], "2020-07-03")

  expect_equal(nrow(resultSacubitrilFirstFemale)[[1]], 1)
  expect_equal(resultSacubitrilFirstFemale$n_persons[1], 1)
  expect_equal(resultSacubitrilFirstFemale$n_events[1], 1)
  expect_equal(resultSacubitrilFirstFemale$incidence_start_date[1], "2020-07-01")

  expect_equal(nrow(resultSacubitrilFirstLaterFemale)[1], 0)


  # get treatment for male subjects
  resultAceInhibitorsMale <- minimiseResults(resultAceInhibitors,
    gender = "Male",
    timeAtRisk = "0 to 30"
  )

  resultArbsMale <- minimiseResults(resultArbs,
    gender = "Male",
    timeAtRisk = "0 to 30"
  )

  resultSacubitrilFirstMale <- minimiseResults(resultSacubitrilFirst,
    gender = "Male",
    timeAtRisk = "0 to 30"
  )

  resultSacubitrilFirstLaterMale <- minimiseResults(resultSacubitrilLater,
    gender = "Male",
    timeAtRisk = "0 to 30"
  )


  expect_equal(nrow(resultAceInhibitorsMale)[[1]], 1)
  expect_equal(resultAceInhibitorsMale$n_persons[1], 1)
  expect_equal(resultAceInhibitorsMale$n_events[1], 1)
  expect_equal(resultAceInhibitorsMale$incidence_start_date[1], "2020-07-04")

  expect_equal(nrow(resultArbsMale)[[1]], 1)
  expect_equal(resultArbsMale$n_persons[1], 1)
  expect_equal(resultArbsMale$n_events[1], 1)
  expect_equal(resultArbsMale$incidence_start_date[1], "2020-07-05")

  expect_equal(nrow(resultSacubitrilFirstLaterMale)[[1]], 1)
  expect_equal(resultSacubitrilFirstLaterMale$n_persons[1], 2)
  expect_equal(resultSacubitrilFirstLaterMale$n_events[1], 2)
  expect_equal(resultSacubitrilFirstLaterMale$incidence_start_date[1], "2020-07-11")

  expect_equal(nrow(resultSacubitrilFirstMale)[1], 0)

  # general population
  resultPathGeneral <- file.path(resultsDir, "incidence.csv")

  resultGeneral <- omopgenerics::importSummarisedResult(resultPathGeneral)
  resultGeneral <- formatIncidence(resultGeneral)

  resultGeneral <- minimiseResults(resultGeneral,
    gender = "Both",
    timeAtRisk = "0 to Inf"
  )

  # heart failre population
  resultPathHf <- file.path(resultsDir, "incidence_hf.csv")

  resultHf <- omopgenerics::importSummarisedResult(resultPathHf)
  resultHf <- formatIncidence(resultHf)

  resultHf <- minimiseResults(resultHf,
    gender = "Both",
    timeAtRisk = "0 to 30"
  )

  expect_equal(nrow(resultGeneral)[[1]], 1)
  expect_equal(resultGeneral$n_persons[1], 5)
  expect_equal(resultGeneral$n_events[1], 5)
  expect_equal(resultGeneral$incidence_start_date[1], "2015-01-01")

  expect_equal(nrow(resultHf)[[1]], 1)
  expect_equal(resultHf$n_persons[1], 5)
  expect_equal(resultHf$n_events[1], 5)
  expect_equal(resultHf$incidence_start_date[1], "2020-07-01")

  unlink(resultsDir, recursive = TRUE)
})

````

## Test scenario
Strata

````
test_that("runStudy results, strata", {

  testPath <- getwd()
  # Read patients from JSON
  cdm <- TestGenerator::patientsCDM(
    pathJson = testPath,
    testName = "testStrata"
  )

  dbName <- "testStrata"

  resultsDir <- file.path(testPath, glue::glue("Results_{dbName}"))

  runStudy(
    conn = attr(cdm, "dbcon"),
    cdmDatabaseSchema = "main",
    resultsDatabaseSchema = "main",
    dbName = dbName,
    tablePrefix = "pre_",
    minCellCount = 0
  )

  # define result file paths
  resultPathAceInhibitors <- file.path(resultsDir, "incidence_hf_treatment_ace_inhibitors.csv")
  resultPathArbs <- file.path(resultsDir, "incidence_hf_treatment_arbs.csv")
  resultPathSacubitrilFirst <- file.path(resultsDir, "incidence_hf_treatment_sacubitril_first.csv")
  resultPathSacubitrilLater <- file.path(resultsDir, "incidence_hf_treatment_sacubitril_later.csv")

  # load and format results
  resultAceInhibitors <- omopgenerics::importSummarisedResult(resultPathAceInhibitors)
  resultAceInhibitors <- formatIncidence(resultAceInhibitors)
  resultArbs <- omopgenerics::importSummarisedResult(resultPathArbs)
  resultArbs <- formatIncidence(resultArbs)
  resultSacubitrilFirst <- omopgenerics::importSummarisedResult(resultPathSacubitrilFirst)
  resultSacubitrilFirst <- formatIncidence(resultSacubitrilFirst)
  resultSacubitrilLater <- omopgenerics::importSummarisedResult(resultPathSacubitrilLater)
  resultSacubitrilLater <- formatIncidence(resultSacubitrilLater)

  # get results for female subjects - test time at risk
  resultAceInhibitorsFemale <- minimiseResults(resultAceInhibitors,
    gender = "Female",
    timeAtRisk = "0 to 30",
    ageGroup = "50 to 79"
  )

  expect_equal(nrow(resultAceInhibitorsFemale)[[1]], 1)
  expect_equal(resultAceInhibitorsFemale$n_persons[1], 1)
  expect_equal(resultAceInhibitorsFemale$n_events[1], 0)
  expect_equal(resultAceInhibitorsFemale$incidence_start_date[1], "2020-07-02")

  resultAceInhibitorsFemale <- minimiseResults(resultAceInhibitors,
    gender = "Female",
    timeAtRisk = "31 to 90",
    ageGroup = "50 to 79"
  )

  expect_equal(nrow(resultAceInhibitorsFemale)[[1]], 1)
  expect_equal(resultAceInhibitorsFemale$n_persons[1], 1)
  expect_equal(resultAceInhibitorsFemale$n_events[1], 1)
  expect_equal(resultAceInhibitorsFemale$incidence_start_date[1], "2020-08-02")


  resultArbsFemale <- minimiseResults(resultArbs,
    gender = "Female",
    timeAtRisk = "181 to 365",
    ageGroup = "50 to 79"
  )

  expect_equal(nrow(resultArbsFemale)[[1]], 1)
  expect_equal(resultArbsFemale$n_persons[1], 1)
  expect_equal(resultArbsFemale$n_events[1], 0)
  expect_equal(resultArbsFemale$incidence_start_date[1], "2020-12-31")


  resultSacubitrilFirstFemale <- minimiseResults(resultSacubitrilFirst,
    gender = "Female",
    timeAtRisk = "0 to 30",
    ageGroup = "50 to 79"
  )

  expect_equal(nrow(resultSacubitrilFirstFemale)[[1]], 1)
  expect_equal(resultSacubitrilFirstFemale$n_persons[1], 1)
  expect_equal(resultSacubitrilFirstFemale$n_events[1], 1)
  expect_equal(resultSacubitrilFirstFemale$incidence_start_date[1], "2020-07-01")


  resultSacubitrilFirstLaterFemale <- minimiseResults(resultSacubitrilLater,
    gender = "Female",
    timeAtRisk = "0 to 30",
    ageGroup = "50 to 79"
  )

  expect_equal(nrow(resultSacubitrilFirstLaterFemale)[[1]], 0)

  # get results for male subjects - test age groups
  resultAceInhibitorsMale <- minimiseResults(resultAceInhibitors,
    gender = "Male",
    timeAtRisk = "0 to 30",
    ageGroup = "18 to 49"
  )

  expect_equal(nrow(resultAceInhibitorsMale)[[1]], 1)
  expect_equal(resultAceInhibitorsMale$n_persons[1], 1)
  expect_equal(resultAceInhibitorsMale$n_events[1], 0)
  expect_equal(resultAceInhibitorsMale$incidence_start_date[1], "2020-07-14")

  resultArbsMale <- minimiseResults(resultArbs,
    gender = "Male",
    timeAtRisk = "0 to 30",
    ageGroup = "18 to 49"
  )

  expect_equal(nrow(resultArbsMale)[[1]], 1)
  expect_equal(resultArbsMale$n_persons[1], 1)
  expect_equal(resultArbsMale$n_events[1], 0)
  expect_equal(resultArbsMale$incidence_start_date[1], "2020-07-15")

  # underage subject must not be included
  resultSacubitrilFirstMale <- minimiseResults(resultSacubitrilFirst,
    gender = "Male",
    timeAtRisk = "0 to 30",
    ageGroup = "18 to 130"
  )

  expect_equal(nrow(resultSacubitrilFirstMale)[[1]], 1)
  expect_equal(resultSacubitrilFirstMale$n_persons[1], 1)
  expect_equal(resultSacubitrilFirstMale$n_events[1], 0)
  expect_equal(resultSacubitrilFirstMale$incidence_start_date[1], "2020-07-13")

  resultSacubitrilFirstMale <- minimiseResults(resultSacubitrilFirst,
    gender = "Male",
    timeAtRisk = "0 to 30",
    ageGroup = "80 to 130"
  )

  expect_equal(nrow(resultSacubitrilFirstMale)[[1]], 1)
  expect_equal(resultSacubitrilFirstMale$n_persons[1], 1)
  expect_equal(resultSacubitrilFirstMale$n_events[1], 0)
  expect_equal(resultSacubitrilFirstMale$incidence_start_date[1], "2020-07-13")


  resultSacubitrilFirstLaterMale <- minimiseResults(resultSacubitrilLater,
    gender = "Male",
    timeAtRisk = "0 to 30",
    ageGroup = "50 to 79"
  )

  expect_equal(nrow(resultSacubitrilFirstLaterMale)[[1]], 1)
  expect_equal(resultSacubitrilFirstLaterMale$n_persons[1], 2)
  expect_equal(resultSacubitrilFirstLaterMale$n_events[1], 2)
  expect_equal(resultSacubitrilFirstLaterMale$incidence_start_date[1], "2020-07-11")

  # test age groups general population
  resultPathGeneral <- file.path(resultsDir, "incidence.csv")

  resultGeneral <- omopgenerics::importSummarisedResult(resultPathGeneral)
  resultGeneral <- formatIncidence(resultGeneral)

  resultGeneralBoth <- minimiseResults(resultGeneral,
    gender = "Both",
    timeAtRisk = "0 to Inf",
    ageGroup = "18 to 49"
  )

  expect_equal(nrow(resultGeneralBoth)[[1]], 1)
  expect_equal(resultGeneralBoth$n_persons[1], 7)

  resultGeneralBoth <- minimiseResults(resultGeneral,
    gender = "Both",
    timeAtRisk = "0 to Inf",
    ageGroup = "50 to 79"
  )

  expect_equal(nrow(resultGeneralBoth)[[1]], 1)
  expect_equal(resultGeneralBoth$n_persons[1], 5)

  resultGeneralBoth <- minimiseResults(resultGeneral,
    gender = "Both",
    timeAtRisk = "0 to Inf",
    ageGroup = "80 to 130"
  )

  expect_equal(nrow(resultGeneralBoth)[[1]], 1)
  expect_equal(resultGeneralBoth$n_persons[1], 1)

  # test heart failure population
  resultPathHf <- file.path(resultsDir, "incidence_hf.csv")

  resultHf <- omopgenerics::importSummarisedResult(resultPathHf)
  resultHf <- formatIncidence(resultHf)

  resultHfBoth <- minimiseResults(resultHf,
    gender = "Both",
    timeAtRisk = "181 to 365",
    ageGroup = "18 to 130"
  )

  expect_equal(nrow(resultHfBoth)[[1]], 1)
  expect_equal(resultHfBoth$n_persons[1], 4) # includes female subject that develops myoclonus more than a year later
  expect_equal(resultHfBoth$n_events[1], 0)

  resultHfBoth <- minimiseResults(resultHf,
    gender = "Both",
    timeAtRisk = "181 to 365",
    ageGroup = "80 to 130"
  )

  expect_equal(nrow(resultHfBoth)[[1]], 1)
  expect_equal(resultHfBoth$n_persons[1], 1)
  expect_equal(resultHfBoth$n_events[1], 0)

  unlink(resultsDir, recursive = TRUE)
})

````
## Test scenario:
### Making sure the study code runs successfully

````
test_that("run study", {
  
  testPath <- getwd()
  # Read patients from JSON
  cdm <- TestGenerator::patientsCDM(
    pathJson = testPath,
    testName = "testData"
  )
  dbName <- "TestData"

  resultsDir <- file.path(testPath, glue::glue("Results_{dbName}"))
  runStudy(
    conn = attr(cdm, "dbcon"),
    cdmDatabaseSchema = "main",
    resultsDatabaseSchema = "main",
    dbName = dbName,
    tablePrefix = "pre_",
    minCellCount = 0,
    useMultipleTimeAtRiskWindows = FALSE
  )

  expect_true(dir.exists(resultsDir))
  result <- list.files(resultsDir)
  expect_length(result, 15)

  expect_equal(sort(result), sort(c(
     "dbinfo.csv", "incidence.csv", "incidence_hf.csv", "incidence_hf_treatment_ace_inhibitors.csv",
    "incidence_hf_treatment_arbs.csv", "incidence_hf_treatment_sacubitril_first.csv", "incidence_hf_treatment_sacubitril_later.csv",
    glue::glue("log_{dbName}.txt"), "obj1GeneralAttrition.csv", "obj1HFAttrition.csv", "obj2_treatment_ace_inhibitors_attrition.csv", 
    "obj2_treatment_arbs_attrition.csv", "obj2_treatment_sacubitril_first_attrition.csv", "obj2_treatment_sacubitril_later_attrition.csv", "sysInfo.txt"
  )))

  unlink(resultsDir, recursive = TRUE)
})

````
## Unit Testing Framework

Makes easier to review the code and ensure that it runs correctly

1. Cohorts/Phenotyping 
2. Study objectives
3. Study execution (runStudy)

## Exercise: ICU Prescriptions

1. Create a test database with a sample of patients
2. Sample dataset P2-C1-014 ICU Prescriptions

## Objective

Characterize hospitalized patients (ICU admission (yes/no)) 

- Hospitalization
- ICU admission
- Who initiated treatment with drug of interest 
- Presence of COVID-19 infection 
- Mechanical ventilation (yes/no)